### 2013-05-30 paper_draft.tex
Added three DTC survey citations and summaries in introduction. Moved crowdsourcing platform part. Added crowdsourcing table. Fixed some minor things.

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

@@

\subsection*{Other crowdsourcing platforms}
The most similar projects to openSNP are the SNPedia and the Personal Genome Project. The focus of the SNPedia is the aggregation and summarization of primary scientific literature on SNPs. The project uses a Wiki to store and display the data collected by volunteers contributing to the project. The data is mainly organized by the unique Rs-ID, as given by dbSNP. If Rs-IDs are missing the identifiers given by the DTC testing companies may be used, similar to the way openSNP stores the data. For individual SNPs pages may list scientific literature and summaries on the found impact can be given. As those pages are largely created manually and not automatically through database access these summaries may not be complete. openSNP utilizes the SNPedia in that it crawls their data for SNPs, the summary of the impact and the magnitude a SNP has. While they offer a page listing download-URLs the SNPedia does not offer any uploading capabilities for genetic data and has no APIs to easily access the different data sets.

The Personal Genome Project (PGP) has its focus on collecting and hosting genetic data as well as phenotypic one. Unlike openSNP they do not offer a completely open enrollment. For each participant of the PGP eligibility has to be established and participants have to give IRB approved informed consent. This allows for an easier re-use of the data, but at the same time makes it impossible for many people to enroll (e.g. non-US citizens). Depending on the specific use one has for the data this might be preferable to the open approach openSNP takes. While the PGP stores genotyping data as well as exome and genome data sets it is currently impossible to access this data through an API, instead data has to be extracted from their database in a highly manual fashion. The annotation database of the PGP is not aimed at delivering specific publications but instead focusses more on specific traits (IS THAT RIGHT?!). Their annotation data is incorporated into openSNP as well.

COMPARE MORE. DO A TABLE. SHOULD THIS MAYBE BE PUT IN THE DISCUSSION INSTEAD OF RESULTS?

@@

@@

\section*{Tables}
%\begin{table}
%\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped. }  
%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
%\hline
%& Mean \emph{Genotyped} & Mean \emph{Not Genotyped} & Mean Difference & Standard Error\\ 
%\hline
%\textbf{Motivation for sharing data in participants who are already genotyped} & & & & \\
%\hline 
%... curious & 3.82 & 2.66 & 1.159 & 0.193 \\ \hline % checked
%... want to help scientists & 4.64 & 4.18  & 0.465 & 0.128 \\ \hline % checked
%... for personal benefits & 3.77 & 3.32  & 0.448 & 0.183 \\ \hline % checked
%\textbf{Motivation for not sharing in participants who are not planning to get genotyped} & & & & \\ \hline
%... fear of discrimination & 3.09 & 4.15  & 1.06 & 0.195 \\ \hline % checked
%... breach of privacy & 3.01 & 3.68  & 0.666 & 0.211 \\ \hline % checked
%... fear of personalized advertising & 3.03 & 3.88  & 0.848 & 0.208 \\ \hline % checked
%... negative consequences for family members & 2.93 & 3.57  & 0.639 & 0.197 \\ \hline 
%\end{tabular}
%\label{tab:motivations1}
%\end{table}
%
%\begin{table}
%\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with participants who would not share their data.}
%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
%\hline
%& Mean {Sharing} & Mean {Not Sharing} & Mean Difference & Standard Error \\ 
%\hline
%\textbf{Motivation for sharing genotypings in participants who would share} & & & & \\
%\hline
%... curiosity & 3.86 & 1.87 & 1.99 & 0.321 \\ \hline % checked
%... want to help science & 4.70 & 3.13 & 1.57 & 0.199 \\ \hline % checked
%... for personal benefits & 3.68 & 2.73 & 0.951 & 0.308 \\ \hline % checked
%\textbf{Motivation for sharing genotypings in participants who would not share} & & & & \\
%\hline
%... fear of discrimination & 3.21 & 4.73 & 1.52 & 0.322 \\ \hline % checked
%... fear of consequences for family members & 2.79 & 3.93 & 1.146 & 0.32 \\ \hline % checked
%... fear of personalized advertising & 3.17 & 4.29 & 1.112 & 0.357 \\ \hline % checked
%\end{tabular}
%\label{tab:motivations2}
%\end{table}
%\begin{table}[!ht]
%\caption{
%\bf{Table title}}
%\begin{tabular}{|c|c|c|}
%table information
%\end{tabular}
%\begin{flushleft}Table caption
%\end{flushleft}
%\label{tab:label}
% \end{table}

>>>>

% for checkmarks
\usepackage{pifont}% http://ctan.org/pkg/pifont
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

@@

@@

\subsection*{Other crowdsourcing platforms}
Projects similar to openSNP are the SNPedia, the Personal Genome Project and PatientsLikeMe.com (see table \ref{tab:platforms} for an overview). The focus of the SNPedia is the aggregation and summary of primary scientific literature on SNPs. The project uses a Wiki to store and display the data collected by volunteers contributing to the project. The data is mainly organized by the unique Rs-ID, as given by dbSNP. If Rs-IDs are missing the identifiers given by the DTC testing companies may be used, similar to the way openSNP stores the data. For individual SNPs pages may list scientific literature and summaries on the found impact can be given. As those pages are largely created manually and not automatically through database access these summaries may not be complete. openSNP utilizes the SNPedia in that it crawls their data for SNPs, the summary of the impact and the magnitude a SNP has. While they offer a page listing download-URLs the SNPedia does not offer any uploading capabilities for genetic data and has no APIs to easily access the different data sets.

The Personal Genome Project (PGP) has its focus on collecting and hosting genetic as well as phenotypic data. Unlike openSNP they do not offer a completely open enrollment. For each participant of the PGP eligibility has to be established and participants have to give IRB approved informed consent. This allows for an easier re-use of the data, but at the same time makes it impossible for many people to enroll (e.g. non-US citizens). Depending on the specific use one has for the data this might be preferable to the open approach openSNP takes. While the PGP stores genotyping data as well as exome and genome data sets it is currently impossible to access this data through an API, instead data has to be manually extracted from their database. The annotation database of the PGP is not aimed at delivering specific publications but instead focuses more on specific traits (IS THAT RIGHT?!). The  annotation data stored by the PGP is incorporated into openSNP as well.

PatientsLikeMe is a community for patients with life-changing illnesses to track and share the development of their illness with other patients with similar illnesses \cite{Wicks2010}.  This helps patients in understanding their diseases better - 72\% of surveyed participants found the site "moderately" or "very helpful", for example when it comes to starting a new medication (37\% found the site helpful), or when it comes to changing the medication (27\%). The data stored in PatientsLikeMe is open to the public and has been shown to be useful for research, for example in Multiple Sclerosis \cite{Bove2013}.

%COMPARE MORE. DO A TABLE. SHOULD THIS MAYBE BE PUT IN THE DISCUSSION INSTEAD OF RESULTS?

% I think discussion is better - we might even change the comparison a bit and put it into introduction, by just comparing SNPedia, PGP etc.

@@

\section*{Tables}
\begin{table}
\caption{Comparison of crowdsourced genetics platforms.}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
Name & Provides annotation & Provides data & Open enrollment & API & License \\ 
\hline
SNPedia & \checkmark & \xmark & N/A & \checkmark & CC-NC-SA 3.0 \\ 
\hline
PGP & \checkmark & \checkmark & \xmark & \xmark & CC-BY 1.0, CC0 \\
\hline
PatientsLikeMe & \xmark & \checkmark & \checkmark & \xmark & CC-BY-SA 3.0 \\
\hline
openSNP & \checkmark & \checkmark & \checkmark & \checkmark & CC-BY 3.0 \\
\hline
\end{tabular}
\label{tab:platforms}
\end{table}
%\begin{table}
%\caption{Differences in terms of motivation to share genotypings with the public in survey-participants who already received a genotyping compared to participants who are not planning to getting genotyped. }  
%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
%\hline
%& Mean \emph{Genotyped} & Mean \emph{Not Genotyped} & Mean Difference & Standard Error\\ 
%\hline
%\textbf{Motivation for sharing data in participants who are already genotyped} & & & & \\
%\hline 
%... curious & 3.82 & 2.66 & 1.159 & 0.193 \\ \hline % checked
%... want to help scientists & 4.64 & 4.18  & 0.465 & 0.128 \\ \hline % checked
%... for personal benefits & 3.77 & 3.32  & 0.448 & 0.183 \\ \hline % checked
%\textbf{Motivation for not sharing in participants who are not planning to get genotyped} & & & & \\ \hline
%... fear of discrimination & 3.09 & 4.15  & 1.06 & 0.195 \\ \hline % checked
%... breach of privacy & 3.01 & 3.68  & 0.666 & 0.211 \\ \hline % checked
%... fear of personalized advertising & 3.03 & 3.88  & 0.848 & 0.208 \\ \hline % checked
%... negative consequences for family members & 2.93 & 3.57  & 0.639 & 0.197 \\ \hline 
%\end{tabular}
%\label{tab:motivations1}
%\end{table}
%
%\begin{table}
%\caption{Differences in terms of motivations to share genotyping-data, comparison between participants who would share their genotyping data with participants who would not share their data.}
%\begin{tabular}{|p{7cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
%\hline
%& Mean {Sharing} & Mean {Not Sharing} & Mean Difference & Standard Error \\ 
%\hline
%\textbf{Motivation for sharing genotypings in participants who would share} & & & & \\
%\hline
%... curiosity & 3.86 & 1.87 & 1.99 & 0.321 \\ \hline % checked
%... want to help science & 4.70 & 3.13 & 1.57 & 0.199 \\ \hline % checked
%... for personal benefits & 3.68 & 2.73 & 0.951 & 0.308 \\ \hline % checked
%\textbf{Motivation for sharing genotypings in participants who would not share} & & & & \\
%\hline
%... fear of discrimination & 3.21 & 4.73 & 1.52 & 0.322 \\ \hline % checked
%... fear of consequences for family members & 2.79 & 3.93 & 1.146 & 0.32 \\ \hline % checked
%... fear of personalized advertising & 3.17 & 4.29 & 1.112 & 0.357 \\ \hline % checked
%\end{tabular}
%\label{tab:motivations2}
%\end{table}
%\begin{table}[!ht]
%\caption{
%\bf{Table title}}
%\begin{tabular}{|c|c|c|}
%table information
%\end{tabular}
%\begin{flushleft}Table caption
%\end{flushleft}
%\label{tab:label}
% \end{table}
